---
title: "Drug vs Alcohol Deaths"
author: "Group 4"
output:
  html_document:
    fig_width: 8
    highlight: tango
    number_sections: no
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---
## Summary

The data seems to contradict what you read in the media:

1.	Drug related deaths are the higher than alcohol related deaths up until the age of 44.
2.	White race related alcohol deaths are the highest.
3.	Disregarding race, drug realted deaths are highest for the younger population and alcohol related deaths are highest for the older population.

A few things to note: Whites in the 50+ age group, may make up the 
largest part of the population. [kaggle: 2013 American Community Survey](https://www.kaggle.com/mchirico/d/census/2013-american-community-survey/population-by-age-and-race)


Also note the following two ICD10codes:

1.	 Icd10Code K703 
   Description: Alcoholic cirrhosis of liver. 
2.	 The 2nd total highest Icd10Code is F102 
   Description: Mental and behavioural disorders due to use of alcohol: Dependence syndrome.




## Running the Analysis

Below taking a look at Alcohol vs. Drug deaths by age.


```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(RSQLite)
library(scales)
library(dplyr)
library(tidyr)
library(knitr)

library(DT)

db <- dbConnect(dbDriver("SQLite"), "input/database.sqlite")

query <-paste("

select 
    a.Age,a.Total Alcohol_Total ,a.White Alcohol_White,
    a.Black Alcohol_Black,a.All_Other  Alcohol_Other,
    g.Total Drug_Total,
    g.White Drug_White,g.Black Drug_Black,g.All_Other Drug_Other
from (
  select 
      Age,count(*) Total ,sum(Race == 1) as White, 
      sum(Race == 2) as Black, sum(Race != 1 and Race != 2) All_Other
   from DeathRecords d,
     (
     select 
       distinct e.DeathRecordId as id 
       from 
         EntityAxisConditions e,
         (
            select * 
            from Icd10Code where Description like '%alcohol%'
 
         ) as c 
        where e.Icd10Code = c.code) as f
    where d.id = f.id and AgeType = 1 and Age < 500
    group by Age
    ) as a, -- Alcohol

(
 select 
     Age,count(*) Total ,sum(Race == 1) as White, 
     sum(Race == 2) as Black, sum(Race != 1 and Race != 2) All_Other
   from DeathRecords d,
       (select 
           distinct e.DeathRecordId as id 
           from 
              EntityAxisConditions e,
              (
 -- Every Firearm discharge, except Legal intervention
        select 
           Code,Description 
           from Icd10Code
           where description like '%drug%' 
              ) as c 
        where e.Icd10Code = c.code) as f
    where d.id = f.id and AgeType = 1 and Age < 500
    group by Age
    ) as g
     where g.Age = a.Age
     -- End of SQL
")

allAlcoholDrug <- dbGetQuery(db, query)

# -- Just Look at Totals
data <- select(allAlcoholDrug, Alcohol_Total,Drug_Total,Age) %>% 
          gather(Group,Deaths,-Age) 

g <- ggplot(dat = data, 
       aes(x=Age, y=Deaths)) + 
      geom_line(aes(colour=Group, group=Group)) + 
      geom_point(aes(colour=Group, shape=Group, group=Group), size=1)+
  ggtitle("Drugs vs. Alcohol Deaths\nby Age")

```
### Graph of Drugs vs. Alcohol

The graph below is interesting. Note the 2nd jump in deaths, after
age 50.

```{r}
# Show plot
g


```


### Race

Let's further subdivide this by race. And this is where
it starts to get interesting.

```{r}

# Let's see where the jump is coming from.

data <- select(allAlcoholDrug, Alcohol_White, Alcohol_Black,
               Drug_White, Drug_Black,Age) %>% 
               gather(Group,Deaths,-Age) 

ggplot(dat = data, 
       aes(x=Age, y=Deaths)) + 
      geom_line(aes(color=Group, group=Group)) + 
      geom_point(aes(color=Group, shape=Group, group=Group), size=1)+
  ggtitle("Drug vs. Alcohol Deaths\nby Age and Race")



```


### Close Look at the Numbers

Let's take a close look at the numbers. It might help
to see this in table form. 

```{r}
# Look at the numbers
datatable(allAlcoholDrug, class = 'compact')
```


```{r}

db <- dbConnect(dbDriver("SQLite"), "input/database.sqlite")

query <-paste("

select *

from (

select Icd10Code,count(*) count from 

( -- Put the whole previous query block in here

 select distinct d.id
   from DeathRecords d,
     (select distinct e.DeathRecordId as id from 
     EntityAxisConditions e,
     (

     -- Every Firearm discharge, except Legal intervention
    select Code,Description from Icd10Code
     where description like '%alcohol%'
 
   ) as c 
  where e.Icd10Code = c.code) as f
  where d.id = f.id and AgeType = 1 

  -- End of big query block
) as s

,DeathRecords d
where s.id=d.id
group by Icd10Code order by count desc


) as a,
Icd10Code b where
a.Icd10Code = b.code limit 10

")

vData <- dbGetQuery(db, query)

```
### Top Icd10Codes

See the table below for an explaination of the top
ICD10codes. Please note the top 2 codes, and the
count (total number of deaths) for each entry.  

The top code K703, list 'Alcoholic cirrhosis of liver'.



```{r}

# This might explain why the results seem odd.
kable(vData)




```



## Explaination:

K703 "Alcoholic cirrhosis of liver" is skewing the results. You can see in the graph below W_K703 (Race White) compared to B_K703 (Race Black) the differences.  



```{r}

db <- dbConnect(dbDriver("SQLite"), "input/database.sqlite")

query <-paste("


select Age,count(*) Total ,sum(Race == 1) as White, 
  sum(Race == 2) as Black, sum(Race != 1 and Race != 2) All_Other,
  sum(Race == 1 and Icd10Code == 'K703')  W_K703,
  sum(Race == 2 and Icd10Code == 'K703')  B_K703,

  sum(Race == 1 and Icd10Code == 'F102')  W_F102,
  sum(Race == 2 and Icd10Code == 'F102')  B_F102,

  sum(Race == 1 and Icd10Code == 'X42')  W_X42,
  sum(Race == 2 and Icd10Code == 'X42')  B_X42,

  sum(Race == 1 and Icd10Code == 'F101')  W_F101,
  sum(Race == 2 and Icd10Code == 'F101')  B_F101,

  sum(Race == 1 and Icd10Code == 'K709')  W_K709,
  sum(Race == 2 and Icd10Code == 'K709')  B_K709,

  sum(Race == 1 and Icd10Code == 'X44')  W_X44,
  sum(Race == 2 and Icd10Code == 'X44')  B_X44





 from DeathRecords d,
   (select distinct e.DeathRecordId as id from 
   EntityAxisConditions e,
   (

   select Code,Description from Icd10Code
   where description like '%alcohol%'
 
   
 
   ) as c 
  where e.Icd10Code = c.code) as f
  where d.id = f.id and AgeType = 1 and Age < 500
 group by Age;

")

v2Data <- dbGetQuery(db, query)

data <- select(v2Data, W_X44,B_X44,
               Age) %>% 
  gather(Group,Total,-Age) 

g <- ggplot(dat = data, 
       aes(x=Age, y=Total)) + 
      geom_line(aes(color=Group, group=Group)) + 
      geom_point(aes(color=Group, shape=Group, group=Group), size=1)+
  ggtitle("K703 - Alcoholic cirrhosis of liver \nby Age")

```
### Graph of K703

Here's the actual graph of K703.  You can see the impact.


```{r}
# Display graph
g

```

### Listing the Table


```{r}


# Here's more detail
datatable(v2Data, class = 'compact')





```





